---
title: "Data Visualization - Summer Olympics"
author: "Kalyani Subbiah"
date: '2019-02-17'
output:
  html_notebook
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
#### 1. Medal Counts over time

```{r}
library(ggplot2)
library(DT)
library(dplyr)
library(plotly)
```

Keep original names because the boundaries of the countries often change with the name changes. For example, the Soviet Union consisted of Russia and several other countries, as did Yugoslavia. Conversely, West Germany and East Germany combined to form Germany. 
```{r}
ath = read.csv("https://raw.githubusercontent.com/qmss-gr5063-2019/course_materials/master/Exercises/03_olympics/athletes_and_events.csv?token=AeYatLKEdXzVZO7Qb6vq5MGD9tMsgut4ks5cfYcIwA%3D%3D")
noc = read.csv("https://raw.githubusercontent.com/qmss-gr5063-2019/course_materials/master/Exercises/03_olympics/noc_regions.csv?token=AeYatOuB_ZcjRQchfHdr4hmwMuBIKnIzks5cfYcxwA%3D%3D")
gdp = read.csv("https://raw.githubusercontent.com/qmss-gr5063-2019/course_materials/master/Exercises/03_olympics/gdp_pop.csv?token=AeYatCJgf39wEd-qOm058uaVbWu-nEByks5cfYciwA%3D%3D")
```

```{r}
master = merge(x = ath, y = noc, by = "NOC", all= TRUE)
gdp$NOC <- gdp$Code
master = merge(x = master, y = gdp, by = "NOC", all= TRUE)
master= unique(master)
``` 
-------
How many summer games did each country commpete in?
```{r}
summer = filter(master, Season == "Summer")
by_year  = summer %>%
  group_by(Year, NOC, add = TRUE)
by_year = by_year %>% select(Year, NOC)
by_year = unique(by_year)
num_summerolympics = table(by_year$NOC)
num_summerolympics = data.frame(num_summerolympics)
names(num_summerolympics) <- c("NOC", "Number")
#sorted by number of olympics they participated in
nums_sorted = num_summerolympics[order(num_summerolympics$Number, decreasing=T),]
datatable(nums_sorted)
```



How many medals of each type did each country win?
```{r}
summer_medal = subset(summer, Medal != 'NA')
summer_medal
by_medal  = summer_medal %>%
  group_by(Medal, NOC)
num_medal = table(by_medal$Medal, by_medal$NOC)
num_medal = data.frame(num_medal)
names(num_medal) <- c("Medal", "NOC", "Number")

#Extracting medal data for 10 countries
NOCs = c("USA", "RUS", "GER","AUS", "CHN",  "IND", "CAN","NOR", "FIN", "AUT")
summer_top = subset(num_medal, NOC %in% NOCs)
summer_top
summer_top$Medal <- factor(summer_top$Medal,levels = c('Gold', 'Silver', 'Bronze'))

medal_plot = ggplot(summer_top, aes(x=NOC, y=Number, fill=Medal, order=Medal)) + geom_bar(stat='identity') + scale_fill_manual(values=c("gold", "grey", "brown"))

medal_plot +  scale_x_discrete(limits=c("USA", "GER","AUS",  "CHN","RUS","CAN","NOR", "FIN", "IND","AUT"), labels = c("United States", "Germany","Australia",  "China","Russia","Canada","Norway", "Finland", "India","Austria")) + theme(axis.text.x = element_text(angle = 90, hjust = 1)) + xlab('') + ylab('Total Number of Medals') + coord_flip() + ggtitle('Total Medals awarded by Country Accross all Summer Olympics')
```


By Gender

```{r}
by_sex  = summer_medal %>%
  group_by(Sex, NOC)
num_sex = table(by_sex$Sex, by_sex$NOC)
num_sex = data.frame(num_sex)
num_sex
names(num_sex) <- c("Sex", "NOC", "Number")
NOCs = c("USA",  "GER","AUS",  "CHN","RUS","CAN","NOR", "FIN", "IND","AUT")
num_sex_top = subset(num_sex, NOC %in% NOCs)
num_sex_top
num_sex_top
sex_plot = ggplot(num_sex_top, aes(x=NOC, y=Number,colour=Sex)) + geom_point() + theme(axis.text.x = element_text(angle = 90, hjust = 1)) + scale_x_discrete(limits=c("USA",  "GER","AUS",  "CHN","RUS","CAN","NOR", "FIN", "IND","AUT"), labels = c("United States", "Germany","Australia",  "China","Russia","Canada","Norway", "Finland", "India","Austria")) + xlab('') + ylab('Total Number of Medals') + coord_flip() + ggtitle('Total Medals awarded by Country and Gender Accross all Summer Olympics')
sex_plot
```
Over Time



```{r}
NOCs = c('USA', 'RUS', 'AUS', 'CHN')

num_year_comp = table(summer_medal$Year, summer_medal$NOC)
num_year_comp
num_year_comp = data.frame(num_year_comp)
num_year_comp
names(num_year_comp) = c("Year", "NOC", "Number")

num_year_comp
num_year_comp = subset(num_year_comp,NOC %in% NOCs)
num_year_comp
num_year_comp$Year = substring(as.Date(num_year_comp$Year, format = "%Y"), 0,4)
num_year_comp
num_year_comp = subset(num_year_comp, Year >= 1984)

year_plot1 = ggplot(num_year_comp, aes(x=Year, y=Number, group = NOC, colour = NOC)) + geom_point() + geom_line()

year_plot1 + theme(axis.text.x = element_text(angle = 90, hjust = 1))   + xlab('') + ylab('Total Number of Medals')  + scale_colour_discrete(guide = guide_legend(reverse=TRUE, title=NULL), labels=c("Australia", "China", "Russia", "United States")) + ggtitle('Total Medals awarded by Country and Year')

```
#### 2. Medal Counts adjusted by Population, GDP

```{r}
num_year_comp = table(summer_medal$Year, summer_medal$Medal, summer_medal$NOC)
num_year_comp = data.frame(num_year_comp)
names(num_year_comp) = c("Year", "Medal", "NOC", "Number")
NOCs = c("USA",  "GER","AUS",  "CHN","RUS","CAN","NOR", "FIN", "IND","AUT")
gold_comp =  subset(num_year_comp,Medal == 'Gold')
gold_comp =  subset(gold_comp,NOC %in% NOCs)
gold_plot = ggplot(gold_comp, aes(x=NOC, y=Number)) + geom_bar(stat='identity', fill = 'gold')
num_year_comp                  
gold_plot + theme(axis.text.x = element_text(angle = 90, hjust = 1)) +   scale_x_discrete(limits=c("USA",  "GER","AUS",  "CHN","RUS","CAN","NOR", "FIN", "IND","AUT"), labels = c("United States", "Germany","Australia",  "China","Russia","Canada","Norway", "Finland", "India","Austria")) + theme(axis.text.x = element_text(angle = 90, hjust = 1)) + xlab('') + ylab('Total Number of Gold Medals') + ggtitle('Total Gold Medals awarded by Country Accross all Summer Olympics')
```

```{r}
summer_medal_gold = subset(summer_medal, Medal == 'Gold')
gold_comp1 <- unique(data.frame(summer_medal_gold$NOC, summer_medal_gold$Population.x))
names(gold_comp1) = c("NOC", "Population")

gold_comp2 <- table(summer_medal$NOC, summer_medal$Medal)
gold_comp2 = data.frame(gold_comp2)
names(gold_comp2) = c("NOC", "Medal", "Number")
gold_comp2= subset(gold_comp2, Medal == 'Gold')
gold_comp1
gold_comp = merge(gold_comp1, gold_comp2, by = 'NOC')

gold_comp <- transform(gold_comp, GoldByPopulation = Number*1000000/(Population))
gold_comp = subset(gold_comp, GoldByPopulation != Inf)
gold_comp = gold_comp[order(gold_comp$GoldByPopulation, decreasing=T),]
gold_comp3 <- unique(data.frame(summer_medal_gold$NOC, summer_medal_gold$region))
gold_comp3$NOC = gold_comp3$summer_medal_gold.NOC

gold_comp = merge(gold_comp, gold_comp3, by = 'NOC')
gold_comp$Country <- gold_comp$summer_medal_gold.region

gold_comp = gold_comp[order(gold_comp$GoldByPopulation, decreasing=T),]
gold_pop = data.frame(gold_comp$Country, gold_comp$GoldByPopulation)
gold_pop = gold_pop[0:15,]
gold_pop$Country = gold_pop$gold_comp.Country
gold_pop$GoldByPopulation = gold_pop$gold_comp.GoldByPopulation
gold_pop_plot = ggplot(gold_pop, aes(x=Country, y=GoldByPopulation)) + geom_bar(stat='identity', fill = 'gold')
gold_pop_plot + theme(axis.text.x = element_text(angle = 90, hjust = 1), axis.title.y=element_text(size=7.5)) +   scale_x_discrete(limits=gold_pop$Country) + xlab('') + ylab('No. of Gold Medals by Population in Millions') + ggtitle('Gold Medals to Population Ratio for Countries accross all Summer olympics')
```
```{r}

gold_comp1 <- unique(data.frame(summer_medal_gold$NOC, summer_medal_gold$GDP.per.Capita.x))
names(gold_comp1) = c("NOC", "GDP")

gold_comp = merge(gold_comp1, gold_comp2, by = 'NOC')
gold_comp
gold_comp <- transform(gold_comp, GoldByGDP = Number*1000/(GDP))
gold_comp = subset(gold_comp, GoldByGDP != Inf)
gold_comp
gold_comp = gold_comp[order(gold_comp$GoldByGDP, decreasing=T),]
gold_comp3 <- unique(data.frame(summer_medal_gold$NOC, summer_medal_gold$region))
gold_comp3$NOC = gold_comp3$summer_medal_gold.NOC
gold_comp3$Country = gold_comp3$summer_medal_gold.region

gold_comp = merge(gold_comp, gold_comp3, by = 'NOC')
gold_comp$Country <- gold_comp$summer_medal_gold
gold_comp = subset(gold_comp, NOC != 'URS')
gold_comp = subset(gold_comp, NOC != 'YUG')

gold_comp = gold_comp[order(gold_comp$GoldByGDP, decreasing=T),]
gold_comp$Country = gold_comp$summer_medal_gold.region
gold_pop = data.frame(gold_comp$Country, gold_comp$GoldByGDP)
gold_pop = gold_pop[0:9,]
gold_pop$Country = gold_pop$gold_comp.Country
gold_pop$GoldByGDP = gold_pop$gold_comp.GoldByGDP
gold_pop_plot = ggplot(gold_pop, aes(x=Country, y=GoldByGDP)) + geom_bar(stat='identity', fill = 'gold')
gold_pop_plot + theme(axis.text.x = element_text(angle = 90, hjust = 1)) +   theme(axis.text.x = element_text(angle = 90, hjust = 1), axis.title.y=element_text(size=7.5)) +   scale_x_discrete(limits=gold_pop$Country) + xlab('') + ylab(' No. of Gold Medals/GDP Per Capita(Thousands)') + ggtitle('Gold Medals to GDP Per Capita Ratio for Countries accross all Summer olympics')
```



#### 3. Host Country Advantage


```{r}
countries <- c("Greece", "France", "United States", "United Kingdom",  "Sweden",  "Belgium", "France",  "Netherlands", "United States", "Germany",  "United Kingdom",  "Finland", "Australia", "Italy", "Japan", "Mexico",  "West Germany", "Canada", "Soviet Union", "United States","South Korea", "Spain", "United States", "Australia", "Greece"   ,"China","United Kingdom",  "Brazil")

host_NOCs <- c("FRG", "FRA", "USA", "GBR",  "SWE",  "BEL", "FRA",  "NED", "USA", "FRG",  "GBR",  "FIN", "AUS", "ITA", "JPN", "MEX",  "FRG", "CAN", "URS", "USA","KOR", "ESP", "USA", "AUS", "GRE"   ,"CHN","GBR",  "BRA")

summer_medal = subset(summer_medal, Year != 1906)
gold = subset(summer_medal, Medal == 'Gold')
gold_year = table(gold$NOC, gold$Year)
gold_year = data.frame(gold_year)
names(gold_year) = c('NOC', 'Year', 'Number')
years= as.factor(unique(gold$Year))
years = sort(unique(years))
years
hosts = data.frame(sort(unique(years)), c("Greece", "France", "United States", "United Kingdom",  "Sweden",  "Belgium", "France",  "Netherlands", "United States", "Germany",  "United Kingdom",  "Finland", "Australia", "Italy", "Japan", "Mexico",  "West Germany", "Canada", "Soviet Union", "United States","South Korea", "Spain", "United States", "Australia", "Greece"   ,"China","United Kingdom",  "Brazil"))
    

num_year_comp = table(gold$Year, gold$NOC)
num_year_comp = data.frame(num_year_comp)
names(num_year_comp) = c("Year", "NOC", "Number")
num_year_comp = subset(num_year_comp, NOC %in% c("FRG", "FRA", "USA", "GBR",  "SWE",  "BEL", "FRA",  "NED", "USA", "FRG",  "GBR",  "FIN", "AUS", "ITA", "JPN", "MEX",  "FRG", "CAN", "URS", "USA","KOR", "ESP", "USA", "AUS", "GRE"   ,"CHN","GBR",  "BRA"))

olympic_hosts <- function(){
  
  host_row1 = subset(num_year_comp, Year == years[0])
  host_row1 = subset(host_row1, NOC == host_NOCs[0])
  for (i in 1:29) {
    
    year <- years[i]
    NOC1 <- host_NOCs[i]
    host_row2 = subset(num_year_comp, Year == year)
    host_row2 = subset(host_row2, NOC == NOC1)
    host_row1 = rbind(host_row1, host_row2)
  }
  return(host_row1)
}

average_golds <- function(){
  averages = c()
  host_row1 = subset(num_year_comp, NOC == host_NOCs[0])
  average = mean(host_row1$Number)
  averages = append(averages, average)
  for (i in 1:29) {
    NOC1 <- host_NOCs[i]
    host_row2 = subset(num_year_comp, NOC == NOC1)
    average = mean(host_row2$Number)
    averages = append(averages, average)
  }
  return(averages)
  }
hosts = olympic_hosts()
hosts
averages = average_golds()
length(averages)
averages = averages[-(0:1)]
averages
averages = averages[-length(averages)]
averages
hosts$average_over_years = averages
hosts
library(ggplot2)
gold_host = ggplot(hosts, aes(x=NOC)) + geom_bar(stat='identity', fill = 'gold', aes(y=Number)) + geom_bar(stat='identity', fill = 'red', aes(y=average_over_years)) + scale_x_discrete( labels = c("Greece: 1896", "France: 1900", "United States: 1904", "United Kingdom: 1908",  "Sweden: 1912",  "Belgium: 1920", "France: 1924",  "Netherlands: 1928", "United States: 1932", "Germany: 1936",  "United Kingdom: 1948",  "Finland: 1952", "Australia: 1956", "Italy: 1960", "Japan: 1964", "Mexico: 1968",  "West Germany: 1972", "Canada: 1976", "Soviet Union: 1980", "United States: 1984","South Korea: 1988", "Spain: 1992", "United States: 1996", "Australia: 2000", "Greece: 2004"   ,"China: 2008","United Kingdom: 2012",  "Brazil: 2016")) + theme(axis.text.x = element_text(angle = 90, hjust = 1)) + xlab('') + ylab('Total Number of Gold Medals won by Host')
gold_host + theme(axis.text.x = element_text(angle = 90, hjust = 1), axis.title.y = element_text(size = 8)) +
ggtitle('Host Country Advantage Accross all Summer Olympics')
```

Note: Red denotes the average medals earned by the host country accross all Summer Olympics. Yellow denotes the number of medals earned by the host country in the Olympics hosted by the country.



#### 4. Most successful athletes


```{r}
by_athletes  = summer_medal %>%
  group_by(Name, Medal, region)
by_athletes = unique(by_athletes)
by_athletes$Country = by_athletes$region
num_athletes = table(by_athletes$Name, by_athletes$Country)
num_athletes = data.frame(num_athletes)
names(num_athletes) <- c("Name", 'Country', "Number")
#sorted by number of medals
nums_sorted = num_athletes[order(num_athletes$Number, decreasing=T),]
nums_sorted = subset(nums_sorted, Number != 0)

nums_sorted1 = nums_sorted[0:30,]
nums_sorted1
athletes_plot = ggplot(nums_sorted1, aes(x=reorder(Name, -Number, reverse =TRUE), y=Number, fill = Country)) + geom_bar(stat='identity')+ xlab('') + ylab('Number of Medals')
athletes_plot + theme(axis.text.x = element_text(angle = 90, hjust = 1)) + coord_flip() + ggtitle('Most Medals won accross all Summer Olympics, by Country(Top 30)')
```


b) Choose one or two additional dimensions among gender, height, weight, sport, discipline, event, year, and country to highlight an interesting pattern in the data.
```{r}
by_athletes  = summer_medal %>%
  group_by(Name, Medal, Year, Sex)
by_athletes = unique(by_athletes)
num_athletes = table(by_athletes$Name, by_athletes$Year,by_athletes$Sex)
num_athletes = data.frame(num_athletes)

names(num_athletes) <- c("Name", 'Year',"Sex", "Number")
nums_sorted = subset(nums_sorted, Number != 0)
#sorted by number of medals
nums_sorted = num_athletes[order(num_athletes$Number, decreasing=T),]
nums_sorted1 = nums_sorted[0:30,]
nums_sorted1$Year = substring(as.Date(nums_sorted1$Year, format = "%Y"), 0,4)
nums_sorted1$NameYear <- paste(nums_sorted1$Name,nums_sorted1$Year)
athletes_plot = ggplot(nums_sorted1, aes(x=reorder(NameYear, -Number), y=Number, fill = Sex)) + geom_bar(stat='identity') + xlab('') + ylab('Number of Medals won in a single year: Top 30')

athletes_plot + theme(axis.text.x = element_text(angle = 90, hjust = 1)) + ggtitle('Highest number of medals won in a single Olympics by Gender (Top 30)')
```

#### 5. Interactivity


```{r}
summer_medal$Country = summer_medal$region
by_athletes  = summer_medal %>%
  group_by(Name, Medal, Year, Country, Sex)
by_athletes = unique(by_athletes)
num_athletes = table(by_athletes$Name, by_athletes$Year)
num_athletes = data.frame(num_athletes)
names(num_athletes) <- c("Name", 'Year',"Number")
Sex_Country = unique(data.frame(summer_medal$Name,summer_medal$Sex, summer_medal$Country, summer_medal$Height, summer_medal$Weight, summer_medal$Age))
names(Sex_Country) <- c("Name", 'Sex',"Country", "Height", "Weight", "Age")
num_athletes = merge(num_athletes, Sex_Country, by = 'Name')
num_athletes$won_medal<-ifelse((num_athletes$Number>0), 1, 0)

num_athletes1 = num_athletes[sample(nrow(num_athletes), 10000), ]

library(ggplot2)
library(plotly)

num_athletes1$won_medal = as.factor(num_athletes1$won_medal)
p <- plot_ly(data = num_athletes1, x = ~Weight, y = ~Height, color= ~won_medal) 
p
```


```{r}
ggplotly(sex_plot)
```


#### 6. Data Table

```{r}
summer_medal1 = data.frame(summer_medal$Name,  summer_medal$Sex, summer_medal$Country, summer_medal$Year, summer_medal$Medal, summer_medal$Event )
summer_medal1
names(summer_medal1) = c("Name", "Sex", "Country", "Year","Medal", "Event")
DT <-  datatable(summer_medal1, class = 'cell-border stripe', rownames =FALSE, options = list(
  pageLength = 10, autoWidth = TRUE
), filter = 'top')

DT
DT:::DT2BSClass('display')
DT:::DT2BSClass(c('compact', 'cell-border'))
DT
```

